Hierarchical Inference of the Intrinsic Ellipticity Distribution of Galaxies during Weak Lensing Analysis

### 2013-04-04 docs/ms.tex
Update ms.tex

Weak gravitational lensing has become the most potential tool to explore
the dark universe. One application of Weak gravitational lensing is
that it can be used to measure the mass profile of single galaxy,galaxy
groups and clusters or large scales. Cosmic shear is the mild distortion
of distant galaxy images due to the bending of light by intervening
matter.It provides one of the most promising methods for constraining
the nature of dark energy (Albrecht et al. 2006; Peacock\&Schneider
2006). While in galaxy--galaxy weaklensing ,tangential shear distortions
of background galaxies around foreground galaxies, allows direct measurement
of the galaxy-DM correlation

The distortion in Weak lensing is typically ~ 0.01,or ~0.001 for
galaxy-galaxy lensing. It's usually more than 10 times smaller than
the intrinsic shape noise,it's also smaller than the distortion caused
be PSF effect which is usually around ~0.05 level. For shear measurement
the main source of uncertainty comes from the intrinsic shape noise
.Since the unlensed galaxy shape can not be observed directly the information
we have about the intrinsic ellipticity is scarce,other than  the assumption that 
the orientational angle should be isotropic. With
this assumption,an estimator can be made by simply averaging galaxies' complex
ellipticity.However by doing this we will not be able to utilize the
information about how the ellipticity are distributed. Simple average
point estimator can't use the other reasonable assumption about intrinsic
shape distribution --galaxies in different sky patches should have
the same shape distibution--which is the natural result of the homogeneous Universe.

A typical procedure of weak lensing analysis  includes ellipticity
measurement,shear inference and mass profile reconstruction. The measurement
of galaxy shapes is the first step of weak gravitational lensing analysis,since
it's the direct observable that relates with tangential shear.However
in this paper we only concern about how to infer shear from galaxy
ellipticity in a hierarchical way.Therefore in our simulation we merely
set up a simple toy model in which we have the observed ellipticity as
data,and the only measurement error we consider is a Gaussian noise
added on the lensed galaxy ellipticity.We also suppose there is a uniform shear
acting on one all the galaxies in sky patch,

Our approach is to use an exceedingly flexible model for intrinsic
shape distribution,and assumes all the galaxies in different patches
abeys such a distribution.Then we can calculate the joint likelihood
of shpape distribution parameters and shears in different sky patches.With
a classical Gibbs MCMC sampler we are able to sample their posteior
distriibution.After marginalizing the shape parameters, we can obtain
the posterior distribution of shear,which we are interested in.This
approach allows us use the information maximumly without introducing
in any wrong assumption on intrinsic shape distribution.

In galaxy--galaxy lensing ,since the distortion is so weak,what people
usually do is stack many lensing signals together and get the mean
shear map,and then infers the mean mass profile of galaxy or galaxy
cluster. However with HB method we can avoid this brute averaging
,instead we can get the posterior distribution of mass profile for
each lensing system - even though it may have very large uncertainty- we
can calculate their statistical property like galaxy-mass correlation
function via this posterior distribution.

@@

and it's inverse formula 
\begin{equation}
\epsilon_{0}=\frac{\epsilon_{\ell}-g}{1-g^{*}\epsilon_{\ell}}
\end{equation}
from Schramm \& Kayser (1995), Seitz \& Schneider (1997), where $\epsilon$
is represented as a complex variable and g, $g^{*}$ are the reduced
shear and its complex conjugate, respectively. In weak lensing region we have 
\begin{equation}
g=\frac{\gamma}{1-k}\approx\gamma
\end{equation}
$\gamma$ is the shear .This is the assumption we use through out
the paper.$\epsilon$ is defined in terms of the major and minor axes and orientation a, b, $\phi$,
respectively, as
\begin{equation}
\epsilon=\frac{a-b}{a+b}e^{2i\phi}
\end{equation}
Due to the isotropicity of Universe ,for the unlensed ellipticity $\epsilon_{0}$ 
we expect $\phi$ has a uniform distribution
from 0 to $\pi$.Then we can expect 
\begin{equation}
<\epsilon_{\ell}>=g
\end{equation}
Therefore $<\epsilon_{\ell}>$ can be adopted as an estimator of g.

However in our hierarchical Bayesian approach we are able to obtain
the posterior distribution of g ,based on the assumption $\phi$ has
a uniform distribution and galaxies in different sky patches have
the same intrinsic shape distribution.

Suppose we know the ellipticity distribution of unlensed galaxies
based on some parasmeters $\alpha$,then we can write the likelihood
of reduced shear and $\alpha$:

@@

The prior of $\alpha$ some how depends on the model we choose for
intrinsic ellipticity distribution .When we use the step function
model for P($\epsilon_{0}$),which we'll describe latter.We find that
we need to use a prior on $\alpha$ . 
\begin{equation}
p(\vec{\alpha}|\xi)=exp(-\xi\sum(\alpha_{i}-\alpha_{i-1})^{2})
\end{equation}

@@

Since $p(\alpha)$ does not explicitly depends on $\vec{g}$ .$p(\alpha|\vec{g})$is
just the pror distribution $p(\vec{\alpha}|\xi)$.Therefore,we have
marginalized likehood 
\begin{equation}
p(\epsilon_{\ell}|\vec{g})=\int p(\epsilon_{\ell}|\vec{g},\alpha)\, p(\vec{\alpha}|\xi)d\alpha
\end{equation}
Given that marginalized likelihood function ,in principle we could
optimize it and gives out a max-likelihood estimator for $\vec{g}$
in different patches simultaneously.

@@

The correctness of posterior distribution of g depends on the intrinsic
ellipticity distribution P($\epsilon_{0}|\alpha$). Due to the limitation
of human's knowledge we are not able to give an analytical form of
$P(\epsilon_{0}|\alpha)$.Therefore we proposed an exceedingly flexible
model for $P(\epsilon_{0}|\alpha)$,which does not requires further
assumption on $P(\epsilon_{0}|\alpha)$.

We divide the 2-D space r$\in$(0,1) into N bins .The probability
of each bin is $p_{i}=e^{\alpha_{i}}$ .$\overrightarrow{\alpha}$
is a N-D parameter vector to be infered,simultaneously with g.

@@

Meanwhile,in order to show the correctness of our method ,we also
set up a simulation in which we know the true $P(|\epsilon_{0}||\alpha)$
is a Beta distribution ,and thus $\alpha$ is the 2-D parameter (p,q)
of the Beta distribution.We will show the posterior samples using
both of these two methods in next section. Note that for either model
we assume P($\epsilon_{0}$) is isotropic.

@@

One thing to note is that the conditional probability of g in ith
patch only explicitly depends on $\alpha$ ,but not g in other patches.Therefore
with $\alpha$ fixed ,the posterior distribution of reduced shear
in ith patch are independent from each other . 
\begin{equation}
p(\vec{g}_{1}....\vec{g}_{n}|\alpha,Data)=p(\vec{g}_{1}|\alpha,Data_{1})....p(\vec{g}_{n}|\alpha,Data_{n})
\end{equation}
Therefore sampling reduced shear g from conditional distribution can
be reduced to 2D sampling problem,which can be done very quickly.

@@

The probability distribution can be visualised by plotting points at each sample
position. An example in the x, y plane is shown for the posterior distribution of
$\vec{g}$ of 64 sky patches in Fig. .8 galaxy shape profiles in each patch is used as the data.
The number density of samples is proportional to the probability density.The figure shows that the two parameters of $\vec{g}$
do not have strong correlation,since the shape of smaples is close to a circle.
We also shows the histogram of g1 in a single sky patch.Not surprisingly it is close to a Gaussian distribution.
If the probability distribution of parameters are perfect Gaussian,their most probable estimates and uncertainty 
is just the mean and standard deviation of the Gaussian distribution.However,in pratical,we find that the distribution 
of reduced shear $\vec{g}$ will be getting closer to Gaussian,with larger number of galaxies used.When we have more than
8 galaxies in each sky patch,we found it's eligible to make such an approximation that to use sample mean and standard 
deviation as most probable estimates and uncertainty.We will justify this approximation with a bias test in the following
sub section.

\subsection{Bias Test}
In this section,we measured the levels of possible systemmatical
bias of this Hierarchical inference method.In principle given the correct model and prior ,
the hierchical Bayesian inference method would have no bias,if propagate the whole likelihood function.
However,in practical,the bias may be introduced by the model we use for the intrinsic shape distribution,
and the prior regularizer of the prefactor $\xi$ as well as the Gaussian approximation we did for the likelihood
function of $\vec{g}$ .Therefore a bias test is necessary to identify how much bias there is. 
Following Heymans et al. (2006), we quantify the systematic bias in
in terms of the multiplicative error m and additive error c on the
true shear $g^{true}$ 
\begin{equation}
\hat{g_{i}}-g_{i}^{true}=m_{i}g_{i}^{true}+c_{i}
\end{equation}

@@

As described in the previous subsection we use the sample mean as the most probable $g_{i}$ and 
the standard deviation as the uncertainty.A standard weighted minimum $\chi^{2}$ linear fit is performed to get the
most-likely value and uncertainty of m and c.

Due to the limitation of galaxy number we are not able to get an exact value for m and c.However in our result 
both the most-like and uncertainty of m are around $10^{-3}$ and for c they are around $10^{-4}$ . Due to the
relatively large uncertainty,it is only an rough estimate of bias level.

@@

\subsection{Measurement Error}
In this section we  discuss how the shape measurement error will
affect the variance and bias of this Hierarchical Bayesian method.
In principle,if the likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$
is known the likelihood $p(\epsilon_{ob}|g,\alpha)$ can be calculated using the convolution as equation ().
However,due to the limitation of computational time we avoid to do this convolution but use $\epsilon_{ob}$
as an estimator of $\epsilon_{\ell}$ .
The likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$ we use in the simulation is a 2D
Gaussian with mean equals $\epsilon_{\ell}$ and $\sigma$ equals 0.05 .
The result of bias test and mean square error test is shown in...

@@

In summary,We have presented a conceptually hierachical inference
approach for weak lensing shear estimation, with carefully calculated 
uncertainty estimates.This method is intrinsically cope with high noise
levels on the input data and naturally allows bad fit to be identified 
from the residuals map. We also have proposed a flexible non-parametric model 
for intrinsic shape distribution allows us to obtain the correct posterior 
distributions, while avoiding making any unsupported assumptions.The Gibbs
sampler allows us to sample the posterior distribution of shear in large number of
sky patches simultaneously and efficiently.

>>>>

Weak gravitational lensing has become the most potential tool to explore
the dark universe. One application of Weak gravitational lensing is
that it can be used to measure the mass profile of single galaxy,galaxy
groups and clusters or large scales. Cosmic shear is the mild distortion
of distant galaxy images due to the bending of light by intervening
matter.It provides one of the most promising methods for constraining
the nature of dark energy (Albrecht et al. 2006; Peacock\&Schneider
2006). While in galaxy--galaxy weaklensing, tangential shear distortions
of background galaxies around foreground galaxies, allows direct measurement
of the galaxy-DM correlation

The distortion in Weak lensing is typically ~ 0.01,or ~0.001 for
galaxy-galaxy lensing. It's usually more than 10 times smaller than
the intrinsic shape noise, it's also smaller than the distortion caused
be PSF effect which is usually around ~0.05 level. For shear measurement
the main source of uncertainty comes from the intrinsic shape noise
.Since the unlensed galaxy shape can not be observed directly the information
we have about the intrinsic ellipticity is scarce,other than  the assumption that 
the orientational angle should be isotropic. With
this assumption,an estimator can be made by simply averaging galaxies' complex
ellipticity.However by doing this we will not be able to utilize the
information about how the ellipticity are distributed. Simple average
point estimator can't use the other reasonable assumption about intrinsic
shape distribution --galaxies in different sky patches should have
the same shape distibution--which is the natural result of the homogeneous Universe.

A typical procedure of weak lensing analysis  includes ellipticity
measurement,shear inference and mass profile reconstruction. The measurement
of galaxy shapes is the first step of weak gravitational lensing analysis,since
it's the direct observable that relates with tangential shear. However
in this paper we only concern about how to infer shear from galaxy
ellipticity in a hierarchical way. Therefore in our simulation we merely
set up a simple toy model in which we have the observed ellipticity as
data,and the only measurement error we consider is a Gaussian noise
added on the lensed galaxy ellipticity.We also suppose there is a uniform shear
acting on one all the galaxies in sky patch.

Our approach is to use an exceedingly flexible model for intrinsic
shape distribution, and assumes all the galaxies in different patches
abeys such a distribution. Then we can calculate the joint likelihood
of shpape distribution parameters and shears in different sky patches.With
a classical Gibbs MCMC sampler we are able to sample their posteior
distriibution.After marginalizing the shape parameters, we can obtain
the posterior distribution of shear,which we are interested in. This
approach allows us use the information maximumly without introducing
in any wrong assumption on intrinsic shape distribution.

In galaxy--galaxy lensing ,since the distortion is so weak,what people
usually do is stack many lensing signals together and get the mean
shear map,and then infers the mean mass profile of galaxy or galaxy
cluster. However with HB method we can avoid this brute averaging
, instead we can get the posterior distribution of mass profile for
each lensing system - even though it may have very large uncertainty- we
can calculate their statistical property like galaxy-mass correlation
function via this posterior distribution.

@@

and it's inverse formula 
\begin{equation}
\epsilon_{0}=\frac{\epsilon_{\ell}-g}{1-g^{*}\epsilon_{\ell}}
\end{equation}
from Schramm \& Kayser (1995), Seitz \& Schneider (1997), where $\epsilon$
is represented as a complex variable and g, $g^{*}$ are the reduced
shear and its complex conjugate, respectively. In weak lensing region we have 
\begin{equation}
g=\frac{\gamma}{1-k}\approx\gamma
\end{equation}
$\gamma$ is the shear. This is the assumption we use through out
the paper.$\epsilon$ is defined in terms of the major and minor axes and orientation a, b, $\phi$,
respectively, as
\begin{equation}
\epsilon=\frac{a-b}{a+b}e^{2i\phi}
\end{equation}
Due to the isotropicity of Universe, for the unlensed ellipticity $\epsilon_{0}$ 
we expect $\phi$ has a uniform distribution
from 0 to $\pi$.Then we can expect 
\begin{equation}
<\epsilon_{\ell}>=g
\end{equation}
Therefore $<\epsilon_{\ell}>$ can be adopted as an estimator of g.

However in our hierarchical Bayesian approach we are able to obtain
the posterior distribution of g, based on the assumption $\phi$ has
a uniform distribution and galaxies in different sky patches have
the same intrinsic shape distribution.

Suppose we know the ellipticity distribution of unlensed galaxies
based on some parasmeters $\alpha$, then we can write the likelihood
of reduced shear and $\alpha$:

@@

The prior of $\alpha$ some how depends on the model we choose for
intrinsic ellipticity distribution. When we use the step function
model for P($\epsilon_{0}$), which we'll describe latter.We find that
we need to use a prior on $\alpha$. 
\begin{equation}
p(\vec{\alpha}|\xi)=exp(-\xi\sum(\alpha_{i}-\alpha_{i-1})^{2})
\end{equation}

@@

Since $p(\alpha)$ does not explicitly depends on $\vec{g}$ .$p(\alpha|\vec{g})$is
just the pror distribution $p(\vec{\alpha}|\xi)$.Therefore, we have
marginalized likehood 
\begin{equation}
p(\epsilon_{\ell}|\vec{g})=\int p(\epsilon_{\ell}|\vec{g},\alpha)\, p(\vec{\alpha}|\xi)d\alpha
\end{equation}
Given that marginalized likelihood function, in principle we could
optimize it and gives out a max-likelihood estimator for $\vec{g}$
in different patches simultaneously.

@@

The correctness of posterior distribution of g depends on the intrinsic
ellipticity distribution P($\epsilon_{0}|\alpha$). Due to the limitation
of human's knowledge we are not able to give an analytical form of
$P(\epsilon_{0}|\alpha)$.Therefore we proposed an exceedingly flexible
model for $P(\epsilon_{0}|\alpha)$, which does not requires further
assumption on $P(\epsilon_{0}|\alpha)$.

We divide the 2-D space r$\in$(0,1) into N bins .The probability
of each bin is $p_{i}=e^{\alpha_{i}}$. $\overrightarrow{\alpha}$
is a N-D parameter vector to be infered, simultaneously with g.

@@

Meanwhile,in order to show the correctness of our method, we also
set up a simulation in which we know the true $P(|\epsilon_{0}||\alpha)$
is a Beta distribution, and thus $\alpha$ is the 2-D parameter (p,q)
of the Beta distribution.We will show the posterior samples using
both of these two methods in next section. Note that for either model
we assume P($\epsilon_{0}$) is isotropic.

@@

One thing to note is that the conditional probability of g in ith
patch only explicitly depends on $\alpha$, but not g in other patches. Therefore
with $\alpha$ fixed ,the posterior distribution of reduced shear
in ith patch are independent from each other. 
\begin{equation}
p(\vec{g}_{1}....\vec{g}_{n}|\alpha,Data)=p(\vec{g}_{1}|\alpha,Data_{1})....p(\vec{g}_{n}|\alpha,Data_{n})
\end{equation}
Therefore sampling reduced shear g from conditional distribution can
be reduced to 2D sampling problem,which can be done very quickly.

@@

The probability distribution can be visualised by plotting points at each sample
position. An example in the x, y plane is shown for the posterior distribution of
$\vec{g}$ of 64 sky patches in Fig. .8 galaxy shape profiles in each patch is used as the data.
The number density of samples is proportional to the probability density.The figure shows that the two parameters of $\vec{g}$
do not have strong correlation,since the shape of smaples is close to a circle.
We also shows the histogram of g1 in a single sky patch. Not surprisingly it is close to a Gaussian distribution.
If the probability distribution of parameters are perfect Gaussian,their most probable estimates and uncertainty 
is just the mean and standard deviation of the Gaussian distribution.However,in pratical,we find that the distribution 
of reduced shear $\vec{g}$ will be getting closer to Gaussian,with larger number of galaxies used.When we have more than
8 galaxies in each sky patch,we found it's eligible to make such an approximation that to use sample mean and standard 
deviation as most probable estimates and uncertainty. We will justify this approximation with a bias test in the following
sub section.

\subsection{Bias Test}
In this section,we measured the levels of possible systemmatical
bias of this Hierarchical inference method.In principle given the correct model and prior ,
the hierchical Bayesian inference method would have no bias, if propagate the whole likelihood function.
However,in practical, the bias may be introduced by the model we use for the intrinsic shape distribution,
and the prior regularizer of the prefactor $\xi$ as well as the Gaussian approximation we did for the likelihood
function of $\vec{g}$. Therefore a bias test is necessary to identify how much bias there is. 
Following Heymans et al. (2006), we quantify the systematic bias in
in terms of the multiplicative error m and additive error c on the
true shear $g^{true}$ 
\begin{equation}
\hat{g_{i}}-g_{i}^{true}=m_{i}g_{i}^{true}+c_{i}
\end{equation}

@@

As described in the previous subsection we use the sample mean as the most probable $g_{i}$ and 
the standard deviation as the uncertainty. A standard weighted minimum $\chi^{2}$ linear fit is performed to get the
most-likely value and uncertainty of m and c.

Due to the limitation of galaxy number we are not able to get an exact value for m and c.However in our result 
both the most-like and uncertainty of m are around $10^{-3}$ and for c they are around $10^{-4}$ . Due to the
relatively large uncertainty, it is only an rough estimate of bias level.

@@

\subsection{Measurement Error}
In this section we  discuss how the shape measurement error will
affect the variance and bias of this Hierarchical Bayesian method.
In principle,if the likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$
is known the likelihood $p(\epsilon_{ob}|g,\alpha)$ can be calculated using the convolution as equation ().
However,due to the limitation of computational time we avoid to do this convolution but use $\epsilon_{ob}$
as an estimator of $\epsilon_{\ell}$.
The likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$ we use in the simulation is a 2D
Gaussian with mean equals $\epsilon_{\ell}$ and $\sigma$ equals 0.05 .
The result of bias test and mean square error test is shown in...

@@

In summary, we have presented a conceptually hierachical inference
approach for weak lensing shear estimation, with carefully calculated 
uncertainty estimates.This method is intrinsically cope with high noise
levels on the input data and naturally allows bad fit to be identified 
from the residuals map. We also have proposed a flexible non-parametric model 
for intrinsic shape distribution allows us to obtain the correct posterior 
distributions, while avoiding making any unsupported assumptions.The Gibbs
sampler allows us to sample the posterior distribution of shear in large number of
sky patches simultaneously and efficiently.
